---
title: "Skinning Cats with R:<br/> Bathymetric & Coastline Maps"
author: "Tom Wainwright <br/> NOAA Fisheries, Newport, OR <br/>"
date: "Presented 13 Jan. 2015, HMSC UseRs meeting"
output:
  ioslides_presentation:
    fig_height: 5
    fig_width: 8
    highlight: tango
  beamer_presentation:
    highlight: tango
---

## A Typical Problem
- We have some field sampling data, with geographic locations.
- We want to plot them on a map, with bathymetry (or topography), coastlines, and other features
- As always with R
    - <div class="red"> There's more than one way to skin a cat. <div/>
    
## Topics Covered:
- A simple example
    - Bathymetric data - where to get it.
    - Simple plotting with spatial analysis packages: *raster*, *sp*, *rgdal*
- Other topics
    - Coastline data, R packages *rworldmap*, *rworldxtra*
    - *marmap* package
    
## For Example
Trawl locations off Washington Coast, from a gear comparison study (3 cruises, two gear types). 

```{r}
hauls <- read.csv("Hauls.csv", header=TRUE, skip=0, as.is=TRUE)
head(hauls,3)
```

##
Need to compute station midpoints, and numeric cruise and gear codes for plotting:

```{r}
hauls$midLat <- with(hauls, (StartLat + EndLat) / 2)
hauls$midLong <- with(hauls, (StartLong + EndLong) / 2)
hauls$cru <- with(hauls, match(Month, c("May","July","June/July")))
hauls$mmed <- with(hauls, match(Excluder, c("Y","N")))
head(hauls,3)
```

## Where to Get Bathymetry

- The most comprehensive source for bathymetry is ...

## Where to Get Bathymetry

 The most comprehensive source for bathymetry is (YOU GUESSED IT!): **NOAA**

> - NOAA National Geographic Data Center at: 
    - http://www.ngdc.noaa.gov/mgg/bathymetry/relief.html

> - Generally, stick with gridded "DEM" data
    - point soundings and acoustic track data are available if you want to go raw
> - First stop: global ETOPO1 data
    - whole globe on a 1-arc-minute grid
> - If that's too coarse for you, use the "Interactive Map Interface" to find finer-scale regional grids

## Downloading Bathymetry

- Find the data set you want.

> - Select a region close to what you want.

> - *DON'T DOWNLOAD IT!* Just copy the download URL as a template.
    - http://maps.ngdc.noaa.gov/viewers/wcs-client/

> - We'll have R do the download, pasting in lats and longs -
    - that way we have a written record of exactly what data we got
    - that way we can modify the script for a different area
    
> - Netiquette: These can be huge datasets: Don't overload servers/networks by repeating a download every time your script runs!

---

- Now we're ready to download the data:

```{r}
minLat <- 46.5; maxLat <- 47.2
minLon <- -124.5; maxLon <- -123.75
url <- paste("http://maps.ngdc.noaa.gov/mapviewer-support/wcs-proxy/",
             "wcs.groovy?filename=etopo1_bedrock.tif&",
             "request=getcoverage&version=1.0.0&service=wcs&",
             "coverage=etopo1_bedrock&CRS=EPSG:4326&format=geotiff&",
             "resx=0.016666666666666667&resy=0.016666666666666667&bbox=",
             minLon, ",", minLat, ",", maxLon, ",", maxLat, sep="")
fname <- "etopo1_bedrock_GH&WB.tif"
if ( ! file.exists(fname)) {
  download.file(url, fname, mode="wb", cacheOK="false")
  }
```

## Working with geoTIFF files
- We downloaded the data as "geoTIFF" file, a specialized format for gridded geographic data.
- So, load a few libraries, then read the file as "raster" data:
```{r results='hide'}
library(sp)      # Spatial analysis and plotting 
library(rgdal)   # Interface to the GDAL libraries
library(raster)  # Raster (gridded image) datasets
bathy <- raster(fname)
```

---

```{r}
plot(bathy)   ## !UGGLLYY!
```

---

```{r}
plot(bathy, col=topo.colors(32))
contour(bathy, levels=c(-200, -100, -50, 0), add=TRUE)
with(hauls, points(midLong, midLat, pch=(21:23)[cru],
      lwd=2, cex=2, col="black", bg=c("grey50",NA)[mmed]))
```

## We can do better (maybe)
- That's still a bit ugly.
    - The axes don't fit the image
    - The zero contour doesn't match the coastline
    - The big 1-arc-second pixels are a bit distracting at this small a scale
- First change, try to find finer-resolution bathymetry
- Second change, use the spplot() function in the *sp* package
- Third change, use a separate coastlines dataset

### Back to the NOAA data catalog
http://www.ngdc.noaa.gov/mgg/bathymetry/relief.html

## New Dataset: crm data
- Coastal Relief Mapping project has fine-resolution (0.3 arc-sec) bathymetry for most of the US coastline.
```{r}
url.hi <- paste("http://maps.ngdc.noaa.gov/mapviewer-support/wcs-proxy/",
             "wcs.groovy?filename=crm.tif&",
             "request=getcoverage&version=1.0.0&service=wcs&",
             "coverage=crm&CRS=EPSG:4326&format=geotiff&",
             "resx=0.000833333333333334&resy=0.000833333333333334&bbox=",
             minLon, ",", minLat, ",", maxLon, ",", maxLat, sep="")
fname.hi <- "crm_WACoast.tif"
if ( ! file.exists(fname.hi)) {
  download.file(url.hi, fname.hi, mode="wb", cacheOK="false")
  }
bathy.hi <- raster(fname.hi)
```

---

```{r}
plot(bathy.hi, col=topo.colors(32))
contour(bathy.hi, levels=c(-200, -100, -50, 0), add=TRUE)
with(hauls, points(midLong, midLat, pch=(21:23)[cru],
      lwd=2, cex=2, col="black", bg=c("grey50",NA)[mmed]))
```

## Using spplot()
- spplot() is the poor-person's GIS system
- But, everything has to be in a "spatial dataframe" with defined coordinates and mapping projection.
- Bathymetry already comes with its own projection data in the geoTIFF, so we just copy that to the hauls data frame:

```{r}
coordinates(hauls) <- c("midLong","midLat")
proj4string(hauls) <- proj4string(bathy)
```

---

- Now, we're ready to build a plot (STRANGE SYNTAX WARNING!)
    - spplot() by itself draws the raster image
    - layers are added to it as a list of "sp.layout" items
    - we save the plot in an object (bplt), then "print" it later
```{r}
bplt <- spplot(bathy.hi, col.regions=topo.colors(64),
               colorkey=FALSE, scales=list(draw=TRUE),
               xlim = c(-124.5, -123.75), ylim=c(46.5, 47.2),
               sp.layout=list(
                 # add contours:
                      list("sp.lines", rasterToContour(bathy.hi, 
                          levels=c(-1000, -100, -50, 0, 1000))),
                 # add trawl locations coded by cruise & gear:
                      list("sp.points", hauls, pch=(21:23)[hauls$cru],
                           lwd=2, cex=2, col="black",
                           fill=c(rgb(0.2,0.2,0.2,0.5),NA)[hauls$mmed])
                 )
)
```

---

```{r}
plot(bplt)
```

## Drawing nice coastlines
- There's a package for that!
    - *rworldmap* for mapping global coastlines, borders, and country-level data 
- But that's pretty coarse resolution
- No worries, there's a package for that, too!
    - *rworldxtra* data(countriesHigh)
```{r}
library(rworldxtra)
data(countriesHigh)
##print(summary(countriesHigh))  ## NOT RUN
```

## Final plot

- Publication-style: black & white.
- Add a filled polygon for the land:
```{r}
bplt <- spplot(bathy.hi, col.regions=NA, colorkey=FALSE,
               scales=list(draw=TRUE),
               xlim = c(-124.5, -123.75), ylim=c(46.5, 47.2),
               sp.layout=list(
                      list("sp.lines", rasterToContour(bathy.hi, 
                          levels=c(-1000, -100, -50, 1000))),
                      list("sp.polygons", countriesHigh, lwd=2, fill="grey"),
                      list("sp.points", hauls, pch=(21:23)[hauls$cru],
                           lwd=2, cex=2, col="black",
                           fill=c(rgb(0.2,0.2,0.2,0.5),NA)[hauls$mmed])
                 )
)
```

---

```{r}
plot(bplt)
```

## Next Challenges
- Add distance scale bar
- Add legend box for point symbols
- Get coastline to overlay a color raster of bathymetry

## *Marmap* package

- My impressions:
    - A specialized package for mapping bathymetry at regional/global scales
    - Can make very pretty maps
    - Some learning curve for specialized plot() arguments
    - Not as flexible as spplot(), but easier
    
- http://cran.r-project.org/web/packages/marmap/index.html
